---
title: "Extending ggplot"
author:
  - E. Sally Chang
date: "2026-04-01"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

# Extending ggplot() using other packages

## Integrating statistical tests using ggpubr()

The strength of `ggplot2` is also its weakness. Using a layered approach
means that it can often take quite a bit of code to reach a publishable
state, even for routine plots. For new users, it is additionally
difficult to remember what elements are needed to adjust various plot
features, as there are hundreds of functions in the `ggplot2` package.
`ggpubr` is a package designed to address some of these problems. [The
‘ggpubr’ package provides some easy-to-use functions for creating and
customizing ‘ggplot2’- based publication ready
plots](https://rpkgs.datanovia.com/ggpubr/).

`ggpubr` allows us to easily incorporate additional information into our
plots (e.g., mean, individual case information, etc.). Where this
package really shines is its ability to easily include the results of
common statistical tests. **As we saw, base R has functions for
calculating statistics, but then you would need to manually add them to
your plot, which is way more annoying.**

Here, I will give you a taste of how this package is useful for
displaying the results of common statistical tests such as those used to
compare means between groups. **We will cover this in much more detail
in Session 4 on Statistical Analyses!**

### Quick ANOVA

Let's look at `myplot` again - what kind of statistical test are we
going to run to compare the means of the plot types?

```{r}
#| warning: false
myplot_cleaned <- myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none") +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)")

myplot_cleaned
```

Let's see if patient age significantly differed between the plot types
using an Analysis of Variance test (ANOVA) - for now, we will assume our
data meets the assumptions of this test. In the future, we will be using
R functions to check whether this is indeed the case!

The main function in `ggpubr` to compare means between two or more
groups is
[`compare_means()`](https://rpkgs.datanovia.com/ggpubr/reference/compare_means.html),
which contains functionalities of t-test, ANOVA, and other options for
comparing means.

Let's try it out for our data:

```{r}
compare_means(hindfoot_length~plot_type,complete_old,method="anova")
```

Notice, this is not nearly as descriptive as using `aov()` (mentioned
briefly in Session 2 and covered in more detail) and this can only
perform a one-way ANOVA. However, it is a convenient function and plays
nicely with `ggpubr` graphing options.

While an ANOVA or Kruskal-Wallis can tell us there are significant
differences between our groups, pairwise testing can then tell us which
groups are different. With `compare_means()` we can use a T-test or
Wilcoxon Rank Sum as a post-hoc for pairwise comparisons.

```{r}
pairwise <-compare_means(hindfoot_length~plot_type,complete_old)
pairwise
```

### **Integrate these results into our visualizations**

If we are just interested in quickly adding the p-value of our ANOVA to
`myplot`, we can use `stat_compare_means()`. You do not need to run
`compare_means()` prior to `stat_compare_means()`. Look how we can
quickly add the overall result to our existing plot!

```{r}
#| warning: false
myplot_cleaned + stat_compare_means(method = "anova")
```

However, if we want to include multiple comparisons and report p-values
adjusted for multiple comparisons, we can add an additional layer from
`stat_pwc` (PWC = pairwise comaparisons):

```{r}
#| warning: false
myplot_cleaned + stat_compare_means(method = "anova") +  stat_pwc( method = "wilcox_test", label = "p.adj.format")
```

This is a little chaotic, so we can try a few different things to make
it less-so:

1.  **Remove any non-significant tests:**

```{r}
#| warning: false
myplot_cleaned + stat_compare_means(method = "anova") +  stat_pwc( method = "wilcox_test", label = "p.adj.format", hide.ns = TRUE)
```

Well, that didn't do much since they are all significant!

2.  **Select which comparisons to keep and adding them "manually":**

-   First, create a list of vectors of comparisons you actually want to
    display (it will be helpful to look at the `pairwise` object). For
    example, Control vs. Spectab exclosure and Control vs. Rodent
    Exclosure:

    ```{r}
    my_comparisons <- list( c("Control", "Spectab exclosure"), c("Control", "Rodent Exclosure"))
    ```

-   Second, apply this list using another layer of
    `stat_compare_means()` , using the same test we used for calculating
    the comparisons above:

```{r}
#| warning: false
myplot_cleaned + stat_compare_means(method = "anova", label.y = 10) +  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test")
```

3.  **Make things visually clearer by adding the significance levels
    instead of the raw numbers** by adjusting the `label` parameter:

```{r}
#| warning: false
myplot_cleaned + stat_compare_means(method = "anova", label.y = 10) +  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label = "p.signif")
```

Those are just some of the ways that we can adjust our plots! *We will
talk way more about the underlying calculations and how significance
levels are calculated in the next session.*

## ggprism package

[`theme_prism()`](https://csdaw.github.io/ggprism/reference/theme_prism.html)
is arguably the most useful (even according to the authors) part of this
package. It is the main tool which is used make ggplots look like they
were made in GraphPad Prism. *There is also a lot of customization you
can do with the following, all using ggplot-like code.*

-   [Color, fill, and shape
    palettes](https://csdaw.github.io/ggprism/articles/colours.html)

-   [Changing Axes](https://csdaw.github.io/ggprism/articles/axes.html)

-   [Adding
    p-values](https://csdaw.github.io/ggprism/articles/pvalues.html)

We will go through some of the options associated with the
`theme_prism()` function. For a change of pace, let's go use the
scatterplot we worked with before:

```{r}
#| warning: false
scatterp <- ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) + geom_point()
scatterp
```

Then we’ll apply
[`theme_prism()`](https://csdaw.github.io/ggprism/reference/theme_prism.html)
and see how it looks just with the default settings. By default the:

-   Fontface for all text is bold except the legend text

-   Font size is 14 pt

-   Legend title is hidden (by setting `legend.title = element_blank()`)

-   Background is white and all text and lines are black

-   Default palette is `"black_and_white"`

```{r}
#| warning: false
scatterp + theme_prism()
```

You can easily bring back the legend title by overriding the
`legend.title` element from inside the `theme()` . Note that this is the
same place that you can edit legend and title elements in regular
`ggplot()`.

```{r}
#| warning: false
scatterp + theme_prism() + theme(legend.title = element_text())
```

The overall theme text size is adjusted with the `base_size` argument,
as with the standard ggplot2 themes such as
[`theme_bw()`](https://ggplot2.tidyverse.org/reference/ggtheme.html).

We can try adjusting it in the code cell below:

```{r}
#| warning: false
scatterp + theme_prism(base_size = ) + theme(legend.title = element_text())
```

By default the `base_line_size` and `base_rect_size` arguments scales
with the `base_size`, but again, these can be overridden.

```{r}
#| warning: false
scatterp + theme_prism(base_size = 14, base_line_size = ) + theme(legend.title = element_text())
```

You can also change the `base_fontface` (e.g. bold, plain, italic) and
`base_family` (crossplatform compatibility fonts e.g. serif, sans, mono,
or specific fonts e.g. Arial).

::: callout-note
## Challenge 1: Adjusting fonts

By looking at help functions and adding different arguments to
`theme_prism()`, figure out how to change font families and
:::

## **Theme palettes**

There are several different colour palettes that are built into
[`theme_prism()`](https://csdaw.github.io/ggprism/reference/theme_prism.html).

```{r}
names(ggprism_data$themes)
```

Try out some of these themes:

```{r}
#| warning: false
scatterp + theme_prism(palette = ) + theme(legend.title = element_text())
```

You can preview a single palette using the
[`preview_theme()`](https://csdaw.github.io/ggprism/reference/preview_theme.html)
function. It shows a sample plot with the theme applied (as well as
[`scale_colour_prism()`](https://csdaw.github.io/ggprism/reference/scale_colour_prism.html)
and
[`scale_fill_prism()`](https://csdaw.github.io/ggprism/reference/scale_fill_prism.html)).
See [this
page](https://csdaw.github.io/ggprism/articles/web-only/palettes.html)
for images of all 43 available
[`theme_prism()`](https://csdaw.github.io/ggprism/reference/theme_prism.html)
palettes.

```{r}
preview_theme("flames")
```

It is good to try matching
[`theme_prism()`](https://csdaw.github.io/ggprism/reference/theme_prism.html)
palettes with their corresponding
[`scale_colour_prism()`](https://csdaw.github.io/ggprism/reference/scale_colour_prism.html)
and/or
[`scale_fill_prism()`](https://csdaw.github.io/ggprism/reference/scale_fill_prism.html)
palettes.

```{r}
scatterp + theme_prism(palette = "") + scale_color_prism(palette = "")
```

::: callout-note
## Realistic Example

If you want to see more complex adjustments you can make, check out
the`ggprism()` guide, including this [really cool real-life
example](https://csdaw.github.io/ggprism/articles/web-only/ex2-wings.html),
re-creating a plot from a paper originally created in GraphPad Prism 7.
You should now have the skills necessary to follow along!
:::

::: callout-note
## Takeaways

-   Packages exist to extend the functionality of `ggplot2` even further

-   They use very similar, if not identical, syntax as the ggplot2
    package, so learning that first will provide a good basis for more
    customization
:::
