---
title: "lesson03_DataViz_Key"
format: html
editor: visual
---

## Lesson 03 - Data Viz Key

## Markdown Document 1

::: callout-tip
# **Exercise 1**

1.  The current axis label text defaults to what we gave as input to `geom_point` (i.e the column headers). We can change this by **adding additional layers** called `xlab()` and `ylab()` for the x- and y-axis, respectively. Add these layers to the current plot such that the x-axis is labeled "Age (days)" and the y-axis is labeled "Mean expression".

2.  Use the `ggtitle` layer to add a plot title of your choice.

3.  Add the following new layer to the code chunk `theme(plot.title=element_text(hjust=0.5))`.

    -   What does it change? **Center the title!**
    -   How many theme() layers can be added to a ggplot code chunk, in your estimation? **Theoretically unlimited!**
:::

```{r}
#| label: added_axes_labels
# Add axes labels to the plot
ggplot(ratdat) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression")
```

```{r}
#| label: add ggtitle
# Add title to the plot
ggplot(ratdat) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age")
```

```{r}
#| label: center_plot_title
# Center plot title
ggplot(ratdat) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age") +
  theme(plot.title=element_text(hjust=0.5))
```

::: callout-note
## Exercise 2

Create another custom function called `linedraw_theme()` that applies the `linedraw` theme to a plot and changes the legend position to the top of the plot.

Then, apply it to `age_vs_means_geom` (or another plot you've created) to test whether it works!
:::

```{r}
linedraw_theme <- function() {
  theme_linedraw() +
  theme(legend.position = "top")
}
```

```{r}
age_vs_means_geom + linedraw_theme()
```

**Exercise 3:**

Generate a boxplot using the data in the `ratdat` dataframe. Create a ggplot2 code chunk with the following instructions, building iteratively to test your code in the code blocks below:

1.  Use the `ggplot()` and `geom_boxplot()` layers to plot the differences in sample means between the Wt and KO genotypes. I've started this for you in the first code cell below!

2.  Use the `fill` *aesthetic* inside our geom to look at differences in sample means between the celltypes within each genotype.

3.  Add a title to your plot using `ggtitle()`

4.  Add labels, “Genotype” for the x-axis and “Mean expression” for the y-axis.

5.  Make the following `theme()` changes. Hint: You may need to use help functions to look at theme() options.

    -   Use the `theme_bw()` function to make the background white.

    -   Change the size of your axes labels to 1.25x larger than the default.

    -   Change the size of your plot title to 1.5x larger than default.

    -   Center the plot title.

```{r}
# Step 1: Create a boxplot where the x-axis is the genotype and the y-axis are the samplemeans
ggplot(data = ratdat, aes(x = genotype, y = samplemeans)) + geom_boxplot()
```

```{r}
# Step 2: Use fill aesthetic to color by celltype
ggplot(data = ratdat, aes(x = genotype, y = samplemeans, fill = celltype)) + geom_boxplot()
```

```{r}
# Step 3:  Add a title to your plot using ggtitle()
ggplot(data = ratdat, aes(x = genotype, y = samplemeans, fill = celltype)) + geom_boxplot() + ggtitle("Genotype differences in average gene expression")
```

```{r}
# Step 4: Add labels, "Genotype" for the x-axis and "Mean expression" for the y-axis.
ggplot(data = ratdat, 
       aes(x = genotype, y = samplemeans, fill = celltype)) + geom_boxplot() + ggtitle("Genotype differences in average gene expression") + 
  xlab("Genotype") +
  ylab("Mean expression") + 
  theme_bw()
```

```{r}
#Step 5: Make the following `theme()` changes:

ggplot(data = ratdat, aes(x = genotype, y = samplemeans, fill = celltype)) +
  geom_boxplot() +
  ggtitle("Genotype differences in average gene expression") +
  xlab("Genotype") +
  ylab("Mean expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title=element_text(size = rel(1.5))) +
  theme(plot.title=element_text(hjust = 0.5))

```

## Markdown Document 2 

::: callout-note
## Exercise 1 : Modifying plots

1.  Try modifying the plot so that the `shape` of the point varies by `sex`. You will set the `shape` the same way you set the `color`.

Do you think this is a good way to represent `sex` with these data?

2.  Now try changing the plot so that the `color` of the points vary by `year`. Do you notice a difference in the color scale compared to changing color by plot type? Why do you think this happened?
:::

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type, shape = sex)) +
  geom_point(alpha = 0.2)
```

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = year, shape = sex)) +
  geom_point(alpha = 0.2)
```

::: callout-note
## Exercise 2: geom_smooth() help

1.  Take a look at the help for the `geom_smooth()` function to figure out what model is used by default to fit a relationship between the two variables
2.  Edit the command above to use a linear model to fit the relationship. How does this change the result?
:::

::: callout-note
## Geom Smooth Help

**Scroll down to `method`: The answer is, it depends on group size!**

Smoothing method (function) to use, accepts either `NULL` or a character vector, e.g. `"lm"`, `"glm"`, `"gam"`, `"loess"` or a function, e.g. `MASS::rlm` or `mgcv::gam`, `stats::lm`, or `stats::loess`. `"auto"` is also accepted for backwards compatibility. It is equivalent to `NULL`.

For `method = NULL` the smoothing method is chosen based on the size of the largest group (across all panels). [`stats::loess()`](http://127.0.0.1:36727/help/library/stats/help/loess) is used for less than 1,000 observations; otherwise [`mgcv::gam()`](http://127.0.0.1:36727/help/library/mgcv/help/gam) is used with `formula = y ~ s(x, bs = "cs")` with `method = "REML"`. Somewhat anecdotally, `loess` gives a better appearance, but is O(N2)O(N2) in memory, so does not work for larger datasets.

If you have fewer than 1,000 observations but want to use the same `gam()` model that `method = NULL` would use, then set `⁠method = "gam", formula = y ~ s(x, bs = "cs")⁠`.
:::

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2) +
  scale_x_log10() + geom_smooth(method = "lm")
```

::: callout-note
## Exercise 3: New Geoms

1.  Violin plots are similar to boxplots- try making one using `plot_type` and `hindfoot_length` as the x and y variables. Remember that all geom functions start with `geom_`, followed by the type of geom.
2.  Make the color of the points and outlines of the violins vary by `plot_type`, and set the fill of the violins to white. Try playing with the order of the layers to see what looks best.
:::

```{r}
ggplot(data = complete_old, 
       mapping = aes(x = plot_type, 
                     y = hindfoot_length,
                     color = plot_type)) +
  geom_jitter(alpha = 0.2) +
  geom_violin(fill = "white")
```

::: callout-note
## Exercise 4: Customizing a plot's fonts

Modify the previous plot by adding a descriptive subtitle. Increase the font size of the plot title and make it bold.

**Hint**: “bold” is referred to as a font “face”
:::

```{r}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), legend.position = "none",
        plot.title = element_text(face = "bold", size = 20)) +
  labs(title = "Rodent size by plot type",
       subtitle = "Long-term dataset from Portal, AZ",
       x = "Plot type",
       y = "Hindfoot length (mm)")
```

::: callout-note
## Exercise 5: Faceting a Scatterplot

1.  Scroll back up to where we were making a scatterplot with the `geom_smooth()` line. Modify this command to facet this plot by `plot_type`. Adjust the number of rows and columns of the facet to make it visually appealing.
2.  Use `ggsave()` to save this as a file called "rodent_size_plots.jpg". Note: You will need to run the plotting command outside of this document!
:::

```{r}
#| warning: false
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2) +
  scale_x_log10() + geom_smooth() + 
   facet_wrap(~plot_type)
```

```{r}
ggsave("rodent_size_plots.jpg")
```

::: callout-note
## Exercise 6: Your Own Plot!

Try making your own plot! You can run `str(complete_old)` or `?complete_old` to explore variables you might use in your new plot. Pick some that might be of the same type (or involve the same visualizations) as your own data might feature in the future.

Here are a couple ideas to get you started:

-   make a histogram of one of the numeric variables

-   try plotting continuous variables using a line graph

If you feel bold, copy and paste or upload your plot into the shared class document so we can all check it out.
:::
