---
title: "lesson03_DataViz_Key"
format: html
editor: visual
---

## Lesson 03 - Data Viz Key

::: callout-tip
# **Exercise 1**

1.  The current axis label text defaults to what we gave as input to `geom_point` (i.e the column headers). We can change this by **adding additional layers** called `xlab()` and `ylab()` for the x- and y-axis, respectively. Add these layers to the current plot such that the x-axis is labeled "Age (days)" and the y-axis is labeled "Mean expression".

2.  Use the `ggtitle` layer to add a plot title of your choice.

3.  Add the following new layer to the code chunk `theme(plot.title=element_text(hjust=0.5))`.

    -   What does it change? **Center the title!**
    -   How many theme() layers can be added to a ggplot code chunk, in your estimation? **Theoretically unlimited!**
:::

```{r}
#| label: added_axes_labels
# Add axes labels to the plot
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression")
```

```{r}
#| label: add ggtitle
# Add title to the plot
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age")
```

```{r}
#| label: center_plot_title
# Center plot title
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5))) +
  xlab("Age (days)") +
  ylab("Mean expression") +
  ggtitle("Mean Expression by Age") +
  theme(plot.title=element_text(hjust=0.5))
```
