---
title: "Plotting and data visualization in base R"
author:
  - E. Sally Chang
date: "2026-04-01"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

# Plotting in Base R

## About our Data Set

### **The dataset**

In our first example dataset, we have collected whole brain samples from
12 mice and want to evaluate expression differences between them. For
ease of plotting, we are going to focus on the **information about the
data** or **metadata**. Our metadata is also stored in a CSV file. In
this file, each row corresponds to a sample and each column contains
some information about each sample.

The first column contains the row names. The next few columns contain
information about our samples that allow us to categorize them. For
example, the second column contains genotype information for each
sample. Each sample is classified in one of two categories: Wt (wild
type) or KO (knockout). We also have some additional information about
sample mean expression data and the age in days of the mice.

**Today we will can read this data in directly from a URL:**

```{r}
new_metadata <- read.csv("https://github.com/esallychang/prism2r-session3-dataviz/raw/refs/heads/main/data/new_metadata.csv", row.names = 1)
```

### Tidy data 

Once you have the data loaded, please check out the structure using any
commands from Base R or Tidyverse that we already know about:

```{r}

```

One thing you should notice is that the data is in "tidy" format, with
each row representing an observation and each column representing a
variable. **This is exactly the format that plotting functions in R
expect and** means the table can be used to generate almost any plot
type without further rearrangements.

## Plots in base R

R has a number of built-in tools for basic graph types such as
histograms, scatter plots, bar charts, boxplots and much
[more](http://www.statmethods.net/graphs/). Rather than going through
all of different types, we will focus on `plot()`, a generic function
for plotting x-y data.

To get a quick view of the different things you can do with `plot`,
let's use the `example()` function:

```{r}
#| label: plot_example
#| eval: FALSE
# Get an example of a plot
example("plot.default")
```

Here, you will find yourself having to press `<Return>` so you can
scroll through the different types of graphs generated by `plot`. Take
note of the different parameters used with each command and how that
affects the aesthetics of the plot.

```{r}
#| label: exit_example
#| eval: false
# Close the example
dev.off() 
```

### Scatterplot

For some hands-on practice we are going to use `plot` to draw a scatter
plot and obtain a graphical view of the relationship between two sets of
continuous numeric data. From our new_metadata file we will take the
`samplemeans` column and plot it against `age_in_days`, to see how mean
expression changes with age.

Now our metadata has all the information to draw a scatterplot. The base
R function to do this is `plot(y ~ x, data)`:

```{r}
#| label: scatterplot_example
# Create Scatterplot
plot(samplemeans ~ age_in_days, data=new_metadata)
```

Each point represents a sample. The values on the y-axis correspond to
the average expression for each sample which is dependent on the x-axis
variable `age_in_days`. This plot is in its simplest form, we can
customize many features of the plot (fonts, colors, axes, titles)
through [graphic
options](http://www.statmethods.net/advgraphs/parameters.html).

For example, let's start by giving our **plot a title and renaming the
axes**. We can do that by simply adding the options `xlab`, `ylab` and
`main` as arguments to the `plot()` function:

```{r}
#| label: add_title_axes
# Add title and axes
plot(samplemeans ~ age_in_days,
     data=new_metadata, 
     main="Expression changes with age",
     xlab="Age (days)",
     ylab="Mean expression")
```

We can also change the **shape of the data point using the `pch`**
option and the **size of the data points using `cex`** (specifying the
amount to magnify relative to the default).

```{r}
#| label: change_point_size_shape
# Change the point size and shape
plot(samplemeans ~ age_in_days,
     data=new_metadata,
     main="Expression changes with age",
     xlab="Age (days)",
     ylab="Mean expression",
     pch="*",
     cex=2.0)
```

We can also add some **color to the data points** on the plot by adding
`col="blue"`. Alternatively, you can sub in any of the default colors or
you can [experiment with other R
packages](http://www.stat.ubc.ca/~jenny/STAT545A/block14_colors.html#basic-color-specification-and-the-default-palette)
to fiddle with better palettes.

We can also add color to **separate the data points by information** in
our data frame. For example, suppose we wanted to the data points
colored by celltype. We would need to specify a vector of colors and
provide the factor by which we are separating samples. The first level
in our factor vector (which by default is assigned alphabetically) would
get assigned the first color that we list. So in this case, blue
corresponds to celltype A samples and green corresponds to celltype B.

```{r}
#| label: add_color_by_celltype
# Factor celltype
new_metadata$celltype <- factor(new_metadata$celltype)

# Plot data with color
plot(samplemeans ~ age_in_days,
     data=new_metadata,
     main="Expression changes with age",
     xlab="Age (days)",
     ylab="Mean expression",
     pch="*",
     cex=2.0,
     col=c("blue", "green")[celltype])
```

The last thing this plot needs is a **figure legend describing the color
scheme**. It would be great if it created one for you by default, but
with R base functions unfortunately it is not that easy. To draw a
legend on the current plot, you need to run a **new function called
`legend()`** and specify the appropriate arguments. The code to do so is
provided below. Don't worry if it seems confusing, we plan on showing
you a much more intuitive way of plotting your data.

```{r}
#| label: add_legend
# Create plot
plot(samplemeans ~ age_in_days, data=new_metadata,
     main="Expression changes with age",
     xlab="Age (days)",
     ylab="Mean expression",
     pch="*",
     cex=2.0,
     col=c("blue", "green")[celltype])

# Add legend
legend("topleft",
       pch="*",
       col=c("blue", "green"),
       c("A", "B"),
       cex=0.8,
       title="Celltype")
```

::: {.callout-note collapse="TRUE"}
# Other Types of Plots in Base R

*NOTE: we will not run these in class, but the code is provided if you
are interested in exploring more on your own.*

## Barplot

Barplots are useful for comparing the distribution of a quantitative
variable (numeric) between groups or categories. A **barplot** would be
much more useful to compare the samplemeans (numeric variable) for each
sample. We can use `barplot` to draw a single bar representing each
sample and the height indicates the average expression level.

```{r}
#| label: barplot_help
# Open help page for barplot
# Note that there is no "data=" argument for barplot()
?barplot
```

Similar to the scatterplot, we can use additional arguments to specify
the aesthetics that we want to change. For example, changing axis
labeling and adding some color.

```{r}
#| label: barplot_example
# Create barplot
barplot(new_metadata$samplemeans,
        names.arg=c(1:12),
        horiz=TRUE,
        col=c("darkblue", "red")[new_metadata$genotype]) 
```

## Histogram

If we are interested in an overall distribution of numerical data, a
**histogram** is what we'd want. To plot a histogram of the data use the
`hist` command:

```{r}
#| label: histogram_example
# Create histogram
hist(new_metadata$samplemeans)
```

Again, there are many options that we can change by modifying the
default parameters. Let's color in the bars, remove the borders and
increase the number of breaks:

```{r}
#| label: histogram_modifcations
# Add modifications to histogram
hist(new_metadata$samplemeans,
     xlab="Mean expression level",
     main="",
     col="darkgrey",
     border=FALSE) 
```

## Boxplot

Using additional sample information from our metadata, we can use plots
to compare values between different factor levels or categories. For
example, we can compare the sample means across celltypes 'typeA' and
'typeB' using a **boxplot**.

```{r}
#| label: boxplot_example
# Boxplot
boxplot(samplemeans~celltype,
        data=new_metadata)
```
:::

# Fundamentals of ggplot

## What is ggplot and why are we learning it?

**`ggplot2`** is a plotting package that provides helpful commands to
create complex plots from data in a data frame. It provides a more
programmatic interface for specifying what variables to plot, how they
are displayed, and general visual properties. Therefore, we only need
minimal changes if the underlying data change or if we decide to change
from a bar plot to a scatter plot. This helps in creating publication
quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** refers to the name of the package itself. When using the
package we use the function **`ggplot()`** to generate the plots, and so
references to using the function will be referred to as **`ggplot()`**
and the package as a whole as **`ggplot2`**

**`ggplot2`** plots work best with data in the ‘long’ format, i.e., a
column for every variable, and a row for every observation.
Well-structured data will save you lots of time when making figures with
**`ggplot2`**

ggplot graphics are built layer by layer by adding new elements. Adding
layers in this fashion allows for extensive flexibility and
customization of plots.

## The ggplot formula:

To build a ggplot, we will use the following basic template that can be
used for different types of plots:

```         
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()) 
```

**Basically, we need to tell ggplot2():**

-   What dataset to be looking at

-   How to map this data to a plot (for example, what columns are we
    plotting on the axes?)

-   What geometry we should use to represent the actual data (points?
    lines? bars?)

-   By adding more statements you can add more customization to your
    plot!

## Motivation:

Today we will go through creating and customizing a few typical plots so
you can get a handle on the `ggplot2()` syntax. Once you are feeling
comfortable with these fundamentals, you will be able to branch out and
learn how to re-create really specific GraphPad Prism plots and get
inspired by what you see in publications.

**Here are some sources of inspiration:**

-   [R Graph Gallery](https://r-graph-gallery.com/ggplot2-package.html)

-   [Re-creating specific barplot in
    R](https://www.rpubs.com/artur_matysik/graphpad_01)

-   [ggPrism ggplot
    extension](https://csdaw.github.io/ggprism/articles/web-only/ex1-dose.html)
    examples

## What are our goals for today?

-   Explain the syntax of ggplot2
-   Apply ggplot2 package to visualize data
-   Save customizations as scripts and functions
-   Create and customize boxplots using ggplot2()
-   **Get you to the point where you could follow some of the cool
    tutorials above!**

## Building plots with `ggplot2`

Let's start by loading the `ggplot2` library:

```{r}
#| label: load_ggplot2
# Load the ggplot2 library
library(ggplot2)
```

The `ggplot()` function is used to **initialize the basic graph
structure**, then we add to it. The basic idea is that you specify
different parts of the plot using additional functions one after the
other and combine them into a "code chunk" using the `+` operator; the
functions in the resulting code chunk are called layers.

Let's start:

```{r}
#| label: ggplot_df
#| eval: false
# What happens when we just provide ggplot with the dataframe?
ggplot(new_metadata) 
```

You get a blank plot, because you need to **specify additional
information to the** ggplot() command. Let's start by adding those
aesthetics such as which variables are going to be mapping to which:

```{r}
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) 
```

**What do you notice about this plot now? What have we added and what is
still missing?**

"Aesthetics" are set with the aes() function and can be set either
nested within `ggplot()` (applies to the whole plot) or within
`geom_point()` (applies only to that layer).

The `aes()` function has many different arguments, and all of those
arguments take columns from the original data frame as input. It can be
used to specify many plot elements including the following:

-   position (i.e., on the x and y axes)
-   color ("outside" color)
-   fill ("inside" color)
-   shape (of points)
-   linetype
-   size

To start, we specified x- and y-axis since `geom_point` requires the
most basic information about a scatterplot, i.e. what you want to plot
on the x and y axes. All of the other plot elements mentioned above are
optional.

### Adding the data via a geometric object!

The **geom (geometric) object** is the layer that specifies what kind of
plot we want to draw. A plot **must have at least one `geom`**; there is
no upper limit. Examples include:

-   points (`geom_point`, `geom_jitter` for scatter plots, dot plots,
    etc)
-   lines (`geom_line`, for time series, trend lines, etc)
-   boxplot (`geom_boxplot`, for, well, boxplots!)

Let's add a "geom" layer to our plot using the `+` operator, and since
we want a scatter plot so we will use `geom_point()`.

```{r}
#| label: geom_point_df
#| eval: false
# Now we provide ggplot with the data frame and the type of geom object that we want for our plot
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + geom_point() 
```

Now that we have the required aesthetics, let's add some extras like
color to the plot. We can **`color` the points on the plot based on the
genotype column** within `aes()`. You will notice that there are a
default set of colors that will be used so we do not have to specify.

Note that the legend has been conveniently plotted for us and that this
call of the `aes` function is *inside* of `geom_point` as these are
options specific to the mapping of the data to the geometric object.

```{r}
#| label: add_color_column
# Adding color to our figure based upon the genotype
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + geom_point(aes(color = genotype)) 
```

Let's try to have both **celltype and genotype represented on the
plot**. To do this we can assign the `shape` argument in `aes()` to the
celltype column, so that each celltype is plotted with a different
shaped data point.

```{r}
#| label: add_shape_column
# Adding different shapes to our data based upon the celltype
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + 
  geom_point(aes(color = genotype, shape=celltype)) 
```

## Iteratively Making Improvements to a Graph

This is a totally useable graph at this point, but if you are like me,
you might immediately notice some improvements that could be made!

The data points are quite small. We can adjust the **size of the data
points** within the `geom_point()` layer since it applies to those
points, but it should **not be within `aes()`** since we are not mapping
it to a column in the input data frame, instead we are just specifying a
number.

```{r}
#| label: increase_size
# Increase the size of the points
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + 
  geom_point(aes(color = genotype, shape=celltype), size=2.25) 

```

### Theme Layers

The labels on the x- and y-axis are also quite small and hard to read.
To change their size, we need to add an additional **theme layer**. The
ggplot2 `theme` system handles non-data plot elements such as:

-   Axis label aesthetics
-   Plot background
-   Facet label backround
-   Legend appearance

There are built-in themes we can use (i.e. `theme_bw()`) that mostly
change the background/foreground colours, by adding it as additional
layer. Or we can adjust specific elements of the current default theme
by adding the `theme()` layer and passing in arguments for the things we
wish to change. Or we can use both.

Let's add a layer using `theme_bw()`.

```{r}
#| label: add_pre_built_theme
# Adding in a built-in theme

ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + 
  geom_point(aes(color = genotype, shape=celltype), size=2.25) +
  theme_bw()

```

**Do the axis labels or the tick labels get any larger by changing
themes?** No, they don't. But, we can add arguments using `theme()` to
change the size of axis labels ourselves. Since we will be adding this
layer "on top", or after `theme_bw()`, any features we change will
override what is set by the `theme_bw()` layer.

Let's **increase the size of both the axes titles to be 1.5 times the
default size.** When modifying the size of text the `rel()` function is
commonly used to specify a change relative to the default.

```{r}
#| label: increase_axis_titles_size
# Increasing the size of the axes titles

ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + 
  geom_point(aes(color = genotype, shape=celltype), size=2.25) +
  theme_bw() + 
  theme(axis.title = element_text(size=rel(1.5)))	
```

You can use the `example("geom_point")` function here to explore a
multitude of different aesthetics and layers that can be added to your
plot. As you scroll through the different plots, take note of how the
code is modified. You can use this with any of the different geometric
object layers available in ggplot2 to learn how you can easily modify
your plots!

::: callout-note
## ggplot resources

RStudio provide this very [helpful
cheatsheet](https://rstudio.github.io/cheatsheets/data-visualization.pdf)
for plotting using `ggplot2`. Different example plots are provided and
the associated code (i.e which `geom` or `theme` to use in the
appropriate situation.) We also encourage you to peruse through this
useful [online reference](https://ggplot2.tidyverse.org/reference/) for
working with ggplot2.

**Bonus**: You can install a
[package](https://r-graph-gallery.com/package/ggthemes.html) called
`ggthemes()` to get access to a variety of themes, such as those used by
the political blog 538 or the Wall Street Journal!
:::

::: callout-tip
# **Exercise 1**

1.  The current axis label text defaults to what we gave as input to
    `geom_point` (i.e the column headers). We can change this by
    **adding additional layers** called `xlab()` and `ylab()` for the x-
    and y-axis, respectively. Add these layers to the current plot such
    that the x-axis is labeled "Age (days)" and the y-axis is labeled
    "Mean expression".

2.  Use the `ggtitle` layer to add a plot title of your choice.

3.  Add the following new layer to the code chunk
    `theme(plot.title=element_text(hjust=0.5))`.

    -   What does it change?
    -   How many theme() layers can be added to a ggplot code chunk, in
        your estimation?
:::
