---
title: ""
author: "E. Sally Chang"
date: "2026-04-02"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

# Customizing plots for large data sets

So, that wasn't super exciting given how small our input data set was.
Let's use a package containing a bigger experimental dataset to learn
how to work with datasets that are more complicated: The `ratdat`
package contains data from the [Portal
Project](https://github.com/weecology/PortalData), which is a long-term
dataset from Portal, Arizona, in the Chihuahuan desert.

To actually take a look at the data, we can use the `View()` function to
open an interactive viewer, which behaves like a simplified version of a
spreadsheet program. It’s a handy function, but somewhat limited when
trying to view large datasets.

We can find out more about the dataset by using the `str()` function to
examine the **str**ucture of the data.

**Let's quickly make a scatterplot:**

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

::: callout-note
## Plot Warnings

You may notice a warning that missing values were removed. If a variable
necessary to make the plot is missing from a given row of data (in this
case, `hindfoot_length` or `weight`), it can’t be plotted. `ggplot2`
just uses a warning message to let us know that some rows couldn’t be
plotted.
:::

## **Changing aesthetics**

Building **`ggplot`** plots is often an iterative process, so we’ll
continue developing the scatter plot we just made. You may have noticed
that parts of our scatter plot have many overlapping points, making it
difficult to see all the data. We can adjust the transparency of the
points using the `alpha` argument, which takes a value between 0 and 1:

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2)
```

We can also change the color of the points:

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2, color = "blue")
```

### Adding a third variable:

Let’s try coloring our points according to the sampling plot type (plot
here refers to the physical area where rodents were sampled and has
nothing to do with making graphs). Since we’re now mapping a variable
(`plot_type`) to a component of the ggplot2 plot (`color`), we need to
put the argument inside `aes()`:

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2)
```

::: callout-note
## Challenge 1: Modifying plots

1.  Try modifying the plot so that the `shape` of the point varies by
    `sex`. You will set the `shape` the same way you set the `color`.

Do you think this is a good way to represent `sex` with these data?

2.  Now try changing the plot so that the `color` of the points vary by
    `year`. Do you notice a difference in the color scale compared to
    changing color by plot type? Why do you think this happened?
:::

## **Changing scales**

The default discrete color scale isn’t always ideal: it isn’t friendly
to viewers with colorblindness and it doesn’t translate well to
grayscale. However, **`ggplot2`** comes with quite a few other color
scales, including the fantastic `viridis` scales, which are designed to
be colorblind and grayscale friendly.

We can change scales by adding `scale_` functions to our plots.
`viridis` functionality is part of the
[viridis](https://r-graph-gallery.com/package/viridis.html) package.

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2) +
  scale_color_viridis_d()
```

Scales don’t just apply to colors- any plot component that you put
inside `aes()` can be modified with `scale_` functions. Just as we
modified the scale used to map `plot_type` to `color`, we can modify the
way that `weight` is mapped to the `x` axis by using the
`scale_x_log10()` function:

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length, color = plot_type)) +
  geom_point(alpha = 0.2) +
  scale_x_log10()
```

One nice thing about `ggplot` and the `tidyverse` in general is that
groups of functions that do similar things are given similar names. Any
function that modifies a `ggplot` scale starts with `scale_`, making it
easier to search for the right function.

## **Adding Statistical Layers**

In addition to re-calculating variables such as, creating out log-scaled
axis, we can fit a simple relationship to the data by adding another
layer, `geom_smooth()`.

```{r}
ggplot(data = complete_old, mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.2) +
  scale_x_log10() + geom_smooth()
```

::: callout-note
## Challenge 2: geom_smooth() help

1.  Take a look at the help for the `geom_smooth()` function to figure
    out what model is used by default to fit a relationship between the
    two variables
2.  Edit the command above to use a linear model to fit the relationship
:::

## **Customizing a Boxplot**

Let’s try making a boxplot, which will have `plot_type` on the x axis
and `hindfoot_length` on the y axis. As a reminder, we can do this by
adding `geom_boxplot()`:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot()
```

### Adjusting Colors:

Just as we colored the points before, we can color our boxplot by
`plot_type` as well:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, color = plot_type)) +
  geom_boxplot()
```

It looks like `color` has only affected the outlines of the boxplot, not
the rectangular portions. This is because the `color` only impacts
1-dimensional parts of a `ggplot`: points and lines. To change the color
of 2-dimensional parts of a plot, we use `fill`:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot()
```

### Wrapping Text:

One thing you may notice is that the axis labels are overlapping each
other, depending on how wide your plot viewer is. One way to help make
them more legible is to **wrap** the text. We can do that by modifying
the **labels** for the `x` axis `scale`.

We use the `scale_x_discrete()` function because we have a discrete
axis, and we modify the `labels` argument. The function
`label_wrap_gen()` will wrap the text of the labels to make them more
legible.

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length, fill = plot_type)) +
  geom_boxplot() +
  scale_x_discrete(labels = label_wrap_gen(width = 10))
```

### Adding geoms: 

While boxplots can be very useful for summarizing data, it is often
helpful to show the raw data as well. With **`ggplot`**, we can easily
add another `geom_` to our plot to show the raw data.

Let’s add `geom_point()` to visualize the raw data. We will modify the
`alpha` argument to help with overplotting.

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  geom_point(alpha = 0.2)
```

Uh oh…now we see an issue with trying to plot using a big dataset! All
our points for a given `x` axis category fall exactly on a line, which
isn’t very useful.

We can shift to using `geom_jitter()`, which will add points with a bit
of random noise added to the positions to prevent this from happening.

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.2)
```

You may have noticed that some of our data points are now appearing on
our plot twice: the outliers are plotted as black points from
`geom_boxplot()`, but they are also plotted with `geom_jitter()`. Since
we don’t want to represent these data multiple times in the same form
(points), we can stop `geom_boxplot()` from plotting them. We do this by
setting the `outlier.shape` argument to `NA`, which means the outliers
don’t have a shape to be plotted.

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.2)
```

### Global mapping vs. Geom Mapping: 

Just as before, we can map `plot_type` to `color` by putting it inside
`aes()`.

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = plot_type), alpha = 0.2)
```

Notice that both the color of the points and the color of the boxplot
lines changed. Any time we specify an `aes()` mapping inside our initial
`ggplot()` function, that mapping will apply to all our `geom`s.

If we want to limit the mapping to a single `geom`, we can put the
mapping into the specific `geom_` function, like this:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = plot_type), alpha = 0.2)
```

Now our points are colored according to `plot_type`, but the boxplots
are all the same color. One thing you might notice is that even with
`alpha = 0.2`, the points obscure parts of the boxplot. This is because
the `geom_point()` layer comes after the `geom_boxplot()` layer, which
means the points are plotted on top of the boxes. To put the boxplots on
top, we switch the order of the layers:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA)
```

Now we have the opposite problem! The white `fill` of the boxplots
completely obscures some of the points. To address this problem, we can
remove the `fill` from the boxplots altogether, leaving only the black
lines. To do this, we set `fill` to `NA`:

```{r}
ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA)
```

::: callout-note
## Challenge 3: Change geoms

1.  Violin plots are similar to boxplots- try making one using
    `plot_type` and `hindfoot_length` as the x and y variables. Remember
    that all geom functions start with `geom_`, followed by the type of
    geom.
2.  For an *extra challenge*, , make the color of the points and
    outlines of the violins vary by `plot_type`, and set the fill of the
    violins to white. Try playing with the order of the layers to see
    what looks best.
:::

## Adding Labels and Modifying Themes

So far we’ve been changing the appearance of parts of our plot related
to our data and the `geom_` functions, but we can also change many of
the non-data components of our plot.

At this point, we are pretty happy with the basic layout of our plot, so
let's **assign** it to a plot to a named **object**.

```{r}
myplot <- ggplot(data = complete_old, mapping = aes(x = plot_type, y = hindfoot_length)) +
  geom_jitter(aes(color = plot_type), alpha = 0.2) +
  geom_boxplot(outlier.shape = NA, fill = NA)

```

We can change the overall appearance using `theme_` functions. Let’s try
a black-and-white theme by adding `theme_bw()` to our plot:

```{r}
myplot + theme_bw()
```

As you can see, a number of parts of the plot have changed. `theme_`
functions usually control many aspects of a plot’s appearance all at
once, for the sake of convenience. To individually change parts of a
plot, we can use the `theme()` function, which can take many different
arguments to change things about the text, grid lines, background color,
and more. Let’s try changing the size of the text on our axis titles. We
can do this by specifying that the `axis.title` should be an
`element_text()` with `size` set to 14.

```{r}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14))
```

Another change we might want to make is to remove the vertical grid
lines. Since our x axis is categorical, those grid lines aren’t useful.
To do this, inside `theme()`, we will change the `panel.grid.major.x` to
an `element_blank()`.

```{r}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        panel.grid.major.x = element_blank())
```

::: callout-note
## Callout: Removing ggplot() elements

You may have noticed that we have used 3 different approaches to getting
rid of something in `ggplot`:

-   `outlier.shape = NA` to remove the outliers from our boxplot

-   `panel.grid.major.x = element_blank()` to remove the x grid lines

-   `legend.position = "none"` to remove our legend

Why are there so many ways to do what seems like the same thing?? This
is a common frustration when working with R, or with any programming
language. There are a couple reasons for it:

1.  Different people contribute to different packages and functions, and
    they may choose to do things differently.

2.  Code may *appear* to be doing the same thing, when the details are
    actually quite different. The inner workings of `ggplot2` are
    actually quite complex, since it turns out making plots is a very
    complicated process! Because of this, things that seem the same
    (removing parts of a plot), may actually be operating on very
    different components or stages of the final plot.

3.  Developing packages is a highly iterative process, and sometimes
    things change. However, changing too much stuff can make old code
    break. Let’s say removing the legend was introduced as a feature of
    `ggplot2`, and then a lot of time passed before someone added the
    feature letting you remove outliers from `geom_boxplot()`. Changing
    the way you remove the legend, so that it’s the same as the boxplot
    approach, could break all of the code written in the meantime, so
    developers may opt to keep the old approach in place.
:::

Our plot is really shaping up now. However, we probably want to make our
axis titles nicer, and perhaps add a main title to the plot. We can do
this using the `labs()` function:

```{r}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none") +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)")
```

We removed our legend from this plot, but you can also change the titles
of various legends using `labs()`. For example,
`labs(color = "Plot type")` would change the title of a color scale
legend to “Plot type”.

::: callout-note
## Challenge 4: Customizing a plot

Modify the previous plot by adding a descriptive subtitle. Increase the
font size of the plot title and make it bold.

**Hint**: “bold” is referred to as a font “face”
:::

## Faceting

One of the most powerful features of **`ggplot`** is the ability to
quickly split a plot into multiple smaller plots based on a categorical
variable, which is called **faceting**.

So far we’ve mapped variables to the x axis, the y axis, and color, but
trying to add a 4th variable becomes difficult. Changing the shape of a
point might work, but only for very few categories, and even then, it
can be hard to tell the differences between the shapes of small points.

Instead of cramming one more variable into a single plot, we will use
the `facet_wrap()` function to generate a series of smaller plots, split
out by `sex`. We also use `ncol` to specify that we want them arranged
in a single column:

```{r}
myplot +
  theme_bw() +
  theme(axis.title = element_text(size = 14), 
        legend.position = "none", 
        panel.grid.major.x = element_blank()) +
  labs(title = "Rodent size by plot type",
       x = "Plot type",
       y = "Hindfoot length (mm)",
       color = "Plot type") +
  facet_wrap(~sex, ncol = 1)
```

::: callout-caution
## Challenge 5: Faceting a Scatterplot

1.  Scroll back up to where we were making a scatterplot with the
    `geom_smooth()` line. Modify this command to facet this plot by.
    Adjust the number of rows and columns of the facet to make it
    visually appealing.
2.  Use `ggsave()` to save this as a file called
    "rodent_size_plots.jpg". Hint: You will need to save the scatterplot
    as an R object first.
:::
