---
title: "Boxplot visualization"
author: "E. Sally Chang"
date: "2026-04-02"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

## Learning Objectives

-   Generate a box plot using ggplot2
-   Customize the look of the boxplot
-   Add additional information to a boxplot to help viewer interpret the
    figure

Introduce geom_jitter() to show the actual points

And by faceting

Annotating box plots:

Adding labels to ggplot() boxplots - what if we wanted to know the names
of the genes?

Adding ggplot boxplots summary stats using custom functions

Highlighting the mean value in a boxplot etc.

Extending with ggpubr():

-   Adding brackets and p-values to R boxplots

Alternatives to boxplot: Violin plots, rain cloud plots

# What is a boxplot? 

A boxplot provides a graphical view of the distribution of data based on
a five number summary:

-   The top and bottom of the box represent the (1) first and (2) third
    quartiles (25th and 75th percentiles, respectively).
-   The line inside the box represents the (3) median (50th percentile).
-   The whiskers extending above and below the box represent the (4)
    maximum, and (5) minimum of a data set.
-   The whiskers of the plot reach the minimum and maximum values that
    are not outliers.

::: callout-note
In this case, **outliers** are determined using the interquartile range
(IQR), which is defined as: Q3 - Q1. Any values that exceeds 1.5 x IQR
below Q1 or above Q3 are considered outliers and are represented as
points above or below the whiskers.
:::

# Putting together a boxplot in ggplot()

In this exercise you will build on what you learned about ggplot()
syntax in the previous scatterplot exercises to create a custom boxplot.
I promise you can do this with the skills you have now!

::: callout-note
**Exercise 1:**

Generate a boxplot using the data in the `new_metadata` dataframe.
Create a ggplot2 code chunk with the following instructions, building
iteratively to test your code in the code blocks below:

1.  Use the `ggplot()` and `geom_boxplot()` layers to plot the
    differences in sample means between the Wt and KO genotypes. I've
    started this for you in the first code cell below!

2.  Use the `fill` *aesthetic* inside our geom to look at differences in
    sample means between the celltypes within each genotype.

3.  Add a title to your plot using `ggtitle()`

4.  Add labels, “Genotype” for the x-axis and “Mean expression” for the
    y-axis.

5.  Make the following `theme()` changes. Hint: You may need to use help
    functions to look at theme() options.

    -   Use the `theme_bw()` function to make the background white.

    -   Change the size of your axes labels to 1.25x larger than the
        default.

    -   Change the size of your plot title to 1.5x larger than default.

    -   Center the plot title.
:::

```{r}
# Step 1: Create a boxplot where the x-axis is the genotype and the y-axis are the samplemeans
ggplot(data = _______, aes(x = ______, y = _______)) + geom_boxplot()
```

```{r}
# Step 2: Use fill aesthetic to color by celltype

```

```{r}
# Step 3:  Add a title to your plot using ggtitle()

```

```{r}
# Step 4: Add labels, “Genotype” for the x-axis and “Mean expression” for the y-axis.
```

```{r}
# Step 5: Make the suggested theme() changes
```

# Changing the order of boxes in your boxplot()

Let’s say you wanted to have the “Wt” boxplots displayed first on the
left side, and “KO” on the right. How might you go about doing this?

To do this, your first question should be - *How does ggplot2 determine
what to place where on the X-axis?*

-   The order of the genotype on the X axis is in alphabetical order.

-   To change it, you need to make sure that the genotype column is a
    factor

-   And, the levels listed in the factor for that column are in the
    order you want on the x-axis

Here is what it looks like to convert the genotype column of the
new_metadata

```{r}
# Factor genotype
new_metadata$genotype <- factor(new_metadata$genotype, c("Wt", "KO"))
```

Now re-run the final code for your boxplot above to see if that changed!
We can also use this as an opportunity to save our boxplot as an R
object to modify in future steps:

```{r}
base_boxplot <- YOUR CODE FROM ABOVE
```

# Changing default colors

::: callout-tip
# **Exercise 3**

You can color the `base_boxplot` differently by using some specific
layers:

1.  Add a new layer `scale_color_manual(values=c("purple","orange"))`.

-   Do you observe a change?

2.  Replace `scale_color_manual(values=c("purple","orange"))` with
    `scale_fill_manual(values=c("purple","orange"))`.

-   Do you observe a change?
:::

::: callout-note
# **Exercise 4**

-   In the scatterplot we drew in class, add a new layer
    `scale_color_manual(values=c("purple","orange"))`.

-   Do you observe a difference?

-   What do you think is the difference between `scale_color_manual()`
    and `scale_fill_manual()`?
:::

If you are interested in learning more, there are whole lessons just on
color customization in

You are not restricted to using colors by writing them out as character
vectors. You have the choice of a lot of colors in R, and you can do so
by using their *hexadecimal code*. For example, "#FF0000" would be red
and "#00FF00" would be green similarly, "#FFFFFF" would be white and
"#000000" would be black. [Click here for more information about color
palettes in
R](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#hexadecimal-color-code-chart)
or you can use this [helpful
cheatsheet](https://www.stat.auckland.ac.nz/~ihaka/downloads/R-colours-a4.pdf)
that lists many popular colors in R.

# **Adding additional geoms to visualize raw data**

## Using our small dataset: 

One of the most powerful aspects of **`ggplot`** is the way we can add
components to a plot in successive layers. While boxplots can be very
useful for summarizing data, it is often helpful to show the raw data as
well. With **`ggplot`**, we can easily add another `geom_` to our plot
to show the raw data.

Let’s add `geom_point()` to visualize the raw data in the code cell
below.

```{r}
base_boxplot + geom_point()
```

## Using a larger test dataset

So, that wasn't super exciting given how small our input data set was.
Let's use a package containing a bigger experimental dataset:

The `ratdat` package contains data from the [Portal
Project](https://github.com/weecology/PortalData), which is a long-term
dataset from Portal, Arizona, in the Chihuahuan desert.

We will be using a dataset called `complete_old`, which contains older
years of survey data. Let’s try to learn a little bit about the data. We
can use a `?` in front of the name of the dataset, which will bring up
the help page for the data.

### R

```         
?complete_old
```

Here we can read descriptions of each variable in our data.

To actually take a look at the data, we can use the `View()` function to
open an interactive viewer, which behaves like a simplified version of a
spreadsheet program. It’s a handy function, but somewhat limited when
trying to view large datasets.

### R

```         
View(complete_old)
```

If you hover over the tab for the interactive `View()`, you can click
the “x” that appears, which will close the tab.

We can find out more about the dataset by using the `str()` function to
examine the **str**ucture of the data.
