---
title: "Boxplot visualization"
author: "E. Sally Chang"
date: "2026-04-02"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

## Learning Objectives

-   Generate a box plot using ggplot2
-   Customize the look of the boxplot
-   Add additional information to a boxplot to help viewer interpret the
    figure

## Generating a Boxplot with ggplot2

A boxplot provides a graphical view of the distribution of data based on
a five number summary:

-   The top and bottom of the box represent the (1) first and (2) third
    quartiles (25th and 75th percentiles, respectively).
-   The line inside the box represents the (3) median (50th percentile).
-   The whiskers extending above and below the box represent the (4)
    maximum, and (5) minimum of a data set.
-   The whiskers of the plot reach the minimum and maximum values that
    are not outliers.

::: callout-note
In this case, **outliers** are determined using the interquartile range
(IQR), which is defined as: Q3 - Q1. Any values that exceeds 1.5 x IQR
below Q1 or above Q3 are considered outliers and are represented as
points above or below the whiskers.
:::

# Outline of the Boxplot

## Adding points to the Boxplot

Introduce geom_jitter() to show the actual points

Adding categorical variables to a boxplot():

Both using colors

And by faceting

Changing the order of groups in a boxplot

Annotating box plots:

Adding labels to ggplot() boxplots - what if we wanted to know the names
of the genes?

Adding ggplot boxplots summary stats using custom functions

Highlighting the mean value in a boxplot etc.

Extending with ggpubr():

-   Adding brackets and p-values to R boxplots

Alternatives to boxplot: Violin plots, rain cloud plots

::: callout-tip
# [**Exercise 1**](14_boxplot_exercise-Answer_key.qmd#exercise-1)

Generate a boxplot using the data in the `new_metadata` dataframe.
Create a ggplot2 code chunk with the following instructions:

1.  Use the `geom_boxplot()` layer to plot the differences in sample
    means between the Wt and KO genotypes.

2.  Use the `fill` *aesthetic* to look at differences in sample means
    between the celltypes within each genotype.

3.  Add a title to your plot.

4.  Add labels, "Genotype" for the x-axis and "Mean expression" for the
    y-axis.

5.  Make the following `theme()` changes:

    -   Use the `theme_bw()` function to make the background white.
    -   Change the size of your axes labels to 1.25x larger than the
        default.
    -   Change the size of your plot title to 1.5x larger than default.
    -   Center the plot title.
:::

**After running the above code the boxplot should look something like
that provided below.**

```{r}
#| label: boxplot_example
#| echo: false
# Boxplot example
ggplot(new_metadata) +
  geom_boxplot(aes(x = genotype, y = samplemeans, fill = celltype)) +
  ggtitle("Genotype differences in Average Gene Expression") +
  xlab("Genotype") +
  ylab("Mean Expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title = element_text(size = rel(1.5))) +
  theme(plot.title = element_text(hjust = 0.5))
```

# Changing the order of genotype on the Boxplot

::: callout-tip
# [**Exercise 2**](14_boxplot_exercise-Answer_key.qmd#exercise-2)

Let's say you wanted to have the "Wt" boxplots displayed first on the
left side, and "KO" on the right. How might you go about doing this?

To do this, your first question should be - *How does ggplot2 determine
what to place where on the X-axis?*

-   The order of the genotype on the X axis is in alphabetical order.
-   To change it, you need to make sure that the genotype column is a
    factor
-   And, the factor levels for that column are in the order you want on
    the x-axis

1.  Factor the `new_metadata$genotype` column without creating any extra
    variables/objects and change the levels to `c("Wt", "KO")`

2.  Re-run the boxplot code chunk you created for the exercise above.
:::

```{r}
#| label: factored_genotype_boxplot
#| echo: false
# Factor genotype
new_metadata$genotype <- factor(new_metadata$genotype, c("Wt", "KO"))

# Create new boxplot
ggplot(new_metadata) +
  geom_boxplot(aes(x = genotype, y = samplemeans, fill = celltype)) +
  ggtitle("Genotype differences in Average Gene Expression") +
  xlab("Genotype") +
  ylab("Mean Expression") +
  theme_bw() +
  theme(axis.title = element_text(size = rel(1.25))) +
  theme(plot.title = element_text(size = rel(1.5))) +
  theme(plot.title = element_text(hjust = 0.5))
```

# Changing default colors

::: callout-tip
# [**Exercise 3**](14_boxplot_exercise-Answer_key.qmd#exercise-3)

You can color the boxplot differently by using some specific layers:

1.  Add a new layer `scale_color_manual(values=c("purple","orange"))`.

-   Do you observe a change?

2.  Replace `scale_color_manual(values=c("purple","orange"))` with
    `scale_fill_manual(values=c("purple","orange"))`.

-   Do you observe a change?
-   In the scatterplot we drew in class, add a new layer
    `scale_color_manual(values=c("purple","orange"))`.
-   Do you observe a difference?
-   What do you think is the difference between `scale_color_manual()`
    and `scale_fill_manual()`?

3.  Back in your boxplot code, change the colors in the
    `scale_fill_manual()` layer to be your 2 favorite colors.

-   Are there any colors that you tried that did not work?
:::

We have [a separate lesson about using color palettes from the package
RColorBrewer](Aside_ggplot2.qmd#customizing-data-point-colors), if you
are interested.

You are not restricted to using colors by writing them out as character
vectors. You have the choice of a lot of colors in R, and you can do so
by using their *hexadecimal code*. For example, "#FF0000" would be red
and "#00FF00" would be green similarly, "#FFFFFF" would be white and
"#000000" would be black. [Click here for more information about color
palettes in
R](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#hexadecimal-color-code-chart)
or you can use this [helpful
cheatsheet](https://www.stat.auckland.ac.nz/~ihaka/downloads/R-colours-a4.pdf)
that lists many popular colors in R.

::: callout-tip
# [**Exercise 4**](14_boxplot_exercise-Answer_key.qmd#exercise-4)

Find the hexadecimal code for your 2 favorite colors (from the final
exercise) and replace the color names with the hexadecimal codes within
the ggplot2 code chunk.
:::
