---
title: "Plotting with the ggplot2 Package"
author:
  - E. Sally Chang
date: "2026-04-01"
license: "CC-BY-4.0"
editor_options: 
  markdown: 
    wrap: 72
---

## What is ggplot and why are we learning it? 

**`ggplot2`** is a plotting package that provides helpful commands to
create complex plots from data in a data frame. It provides a more
programmatic interface for specifying what variables to plot, how they
are displayed, and general visual properties. Therefore, we only need
minimal changes if the underlying data change or if we decide to change
from a bar plot to a scatter plot. This helps in creating publication
quality plots with minimal amounts of adjustments and tweaking.

**`ggplot2`** refers to the name of the package itself. When using the
package we use the function **`ggplot()`** to generate the plots, and so
references to using the function will be referred to as **`ggplot()`**
and the package as a whole as **`ggplot2`**

**`ggplot2`** plots work best with data in the ‘long’ format, i.e., a
column for every variable, and a row for every observation.
Well-structured data will save you lots of time when making figures with
**`ggplot2`**

ggplot graphics are built layer by layer by adding new elements. Adding
layers in this fashion allows for extensive flexibility and
customization of plots.

## The ggplot formula: 

To build a ggplot, we will use the following basic template that can be
used for different types of plots:

```         
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) +  <GEOM_FUNCTION>()) 
```

**Basically, we need to tell ggplot2():**

-   What dataset to be looking at

-   How to map this data to a plot (for example, what columns are we
    plotting on the axes?)

-   What geometry we should use to represent the actual data (points?
    lines? bars?)

-   By adding more statements you can add more customization to your
    plot!

## Motivation: 

Today we will go through creating and customizing a few typical plots so
you can get a handle on the ggplot2() syntax. Once you are feeling
comfortable with these fundamentals, you will be able to branch out and
learn how to re-create really specific GraphPad Prism plots and get
inspired by what you see in publications.

**Here are some sources of inspiration:**

-   [R Graph Gallery](https://r-graph-gallery.com/ggplot2-package.html)

-   [Re-creating specific barplot in
    R](https://www.rpubs.com/artur_matysik/graphpad_01)

-   [ggPrism ggplot
    extension](https://csdaw.github.io/ggprism/articles/web-only/ex1-dose.html)
    examples

## What are our goals for today? 

-   Explain the syntax of ggplot2
-   Apply ggplot2 package to visualize data
-   Save customizations as scripts and functions
-   Create and customize boxplots using ggplot2()
-   Get you to the point where you could follow some of the cool
    tutorials above!

## Building plots with `ggplot2`

Let's start by loading the `ggplot2` library:

```{r}
#| label: load_ggplot2
# Load the ggplot2 library
library(ggplot2)
```

The `ggplot()` function is used to **initialize the basic graph
structure**, then we add to it. The basic idea is that you specify
different parts of the plot using additional functions one after the
other and combine them into a "code chunk" using the `+` operator; the
functions in the resulting code chunk are called layers.

Let's start:

```{r}
#| label: ggplot_df
#| eval: false
# What happens when we just provide ggplot with the dataframe?
ggplot(new_metadata) 
```

You get a blank plot, because you need to **specify additional
information to the** ggplot() command. Let's start by adding those
aesthetics such as which variables are going to be mapping to which:

```{r}
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) 
```

**What do you notice about this plot now? What have we added and what is
still missing?**

"Aesthetics" are set with the aes() function and can be set either
nested within `ggplot()` (applies to the whole plot) or within
`geom_point()` (applies only to that layer).

The `aes()` function has many different arguments, and all of those
arguments take columns from the original data frame as input. It can be
used to specify many plot elements including the following:

-   position (i.e., on the x and y axes)
-   color ("outside" color)
-   fill ("inside" color)
-   shape (of points)
-   linetype
-   size

To start, we specified x- and y-axis since `geom_point` requires the
most basic information about a scatterplot, i.e. what you want to plot
on the x and y axes. All of the other plot elements mentioned above are
optional.

### Adding the actual geometric object! 

The **geom (geometric) object** is the layer that specifies what kind of
plot we want to draw. A plot **must have at least one `geom`**; there is
no upper limit. Examples include:

-   points (`geom_point`, `geom_jitter` for scatter plots, dot plots,
    etc)
-   lines (`geom_line`, for time series, trend lines, etc)
-   boxplot (`geom_boxplot`, for, well, boxplots!)

Let's add a "geom" layer to our plot using the `+` operator, and since
we want a scatter plot so we will use `geom_point()`.

```{r}
#| label: geom_point_df
#| eval: false
# Now we provide ggplot with the data frame and the type of geom object that we want for our plot
ggplot(data = new_metadata, aes(x = age_in_days, y = samplemeans)) + geom_point() 
```

Now that we have the required aesthetics, let's add some extras like
color to the plot. We can **`color` the points on the plot based on the
genotype column** within `aes()`. You will notice that there are a
default set of colors that will be used so we do not have to specify.
Note that the legend has been conveniently plotted for us.

```{r}
#| label: add_color_column
# Adding color to our figure based upon the genotype

```

Let's try to have both **celltype and genotype represented on the
plot**. To do this we can assign the `shape` argument in `aes()` the
celltype column, so that each celltype is plotted with a different
shaped data point.

```{r}
#| label: add_shape_column
# Adding different shapes to our data based upon the celltype
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype)) 
```

The data points are quite small. We can adjust the **size of the data
points** within the `geom_point()` layer, but it should **not be within
`aes()`** since we are not mapping it to a column in the input data
frame, instead we are just specifying a number.

```{r}
#| label: increase_size
# Increase the size of the points
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) 
```

The labels on the x- and y-axis are also quite small and hard to read.
To change their size, we need to add an additional **theme layer**. The
ggplot2 `theme` system handles non-data plot elements such as:

-   Axis label aesthetics
-   Plot background
-   Facet label backround
-   Legend appearance

There are built-in themes we can use (i.e. `theme_bw()`) that mostly
change the background/foreground colours, by adding it as additional
layer. Or we can adjust specific elements of the current default theme
by adding the `theme()` layer and passing in arguments for the things we
wish to change. Or we can use both.

Let's add a layer `theme_bw()`.

```{r}
#| label: add_pre_built_theme
# Adding in a built-in theme
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=3.0) +
  theme_bw() 
```

Do the axis labels or the tick labels get any larger by changing themes?

No, they don't. But, we can add arguments using `theme()` to change the
size of axis labels ourselves. Since we will be adding this layer "on
top", or after `theme_bw()`, any features we change will override what
is set by the `theme_bw()` layer.

Let's **increase the size of both the axes titles to be 1.5 times the
default size.** When modifying the size of text the `rel()` function is
commonly used to specify a change relative to the default.

```{r}
#| label: increase_axis_titles_size
# Increasing the size of the axes titles
ggplot(new_metadata) +
  geom_point(aes(x = age_in_days, y= samplemeans, color = genotype,
  			shape=celltype), size=2.25) +
  theme_bw() +
  theme(axis.title = element_text(size=rel(1.5)))			
```

::: callout-note
You can use the `example("geom_point")` function here to explore a
multitude of different aesthetics and layers that can be added to your
plot. As you scroll through the different plots, take note of how the
code is modified. You can use this with any of the different geometric
object layers available in ggplot2 to learn how you can easily modify
your plots!
:::

::: callout-note
RStudio provide this very [useful
cheatsheet](cheatsheets/data-visualization-2.1.pdf) for plotting using
`ggplot2`. Different example plots are provided and the associated code
(i.e which `geom` or `theme` to use in the appropriate situation.) We
also encourage you to peruse through this useful [online
reference](https://ggplot2.tidyverse.org/reference/) for working with
ggplot2.
:::

::: callout-tip
# [**Exercise 1**](12_ggplot2-Answer_key.qmd#exercise-1)

1.  The current axis label text defaults to what we gave as input to
    `geom_point` (i.e the column headers). We can change this by
    **adding additional layers** called `xlab()` and `ylab()` for the x-
    and y-axis, respectively. Add these layers to the current plot such
    that the x-axis is labeled "Age (days)" and the y-axis is labeled
    "Mean expression".

2.  Use the `ggtitle` layer to add a plot title of your choice.

3.  Add the following new layer to the code chunk
    `theme(plot.title=element_text(hjust=0.5))`.

    -   What does it change?
    -   How many theme() layers can be added to a ggplot code chunk, in
        your estimation?
:::
